Retrofitting Effect Handlers onto OCaml

### 2020-11-19 retro-concurrency.tex
Updates

The biggest impact was the increase in the code size due to stack overflow
checks. We define code size as the sum of the sizes of all the OCaml text
sections in the compiled binary file. The code size metric excludes data, debug
symbols, the runtime (implemented in C) and other C libraries. Compared to
|stock|, the code size for OCaml code is 19\% more for |MC| and |MC+RedZone32|,
and 30\% more for |MC+RedZone0|. The result shows that our 16-word red zone is
effective at reducing the code size compared to having no red zone, whereas the
32-word red zone does not noticeably reduce the code size further. Further work
is required to bring the code size closer to |stock|.

@@

The results show that exceptions are no more expensive in |MC| compared to
|stock|. In the other programs, |MC| executes more instructions due to stack
overflow checks. External calls and callbacks are more expensive due to
additional bookkeeping. Callbacks are significantly more expensive in |MC|.
Since we reuse the current fiber stack for callbacks, we need to ensure it has
enough room for inserting additional frames, while |stock| does not need to do
this. Callback performance is less important than external calls, which are far
more numerous. As seen previously, these results however do not affect macro
benchmark performance.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. The
code does not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

@@

\begin{figure*}
	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml code size of macro benchmarks. Baseline is Stock
		OCaml, whose size in bytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
macro benchmarks that do not use effects using the Multicore OCaml compiler
which supports effect handlers. The aim is to quantify the impact of the
addition of effect handlers to the language on code that does not use effect
handlers. We define code size as the sum of the sizes of all the text sections
in the binary generated by compiling the OCaml code. Thus, the code size metric
excludes data, debug symbols, the runtime (implemented in C) and other
statically linked C libraries. The |MC| is Multicore OCaml with the default red
zone size (16 words). |MC+RedZone0| uses red zone size of 0; every OCaml
function has a stack overflow check. |MC+RedZone32| uses a red zone size of 32
words.

>>>>

The biggest impact was the increase in the OCaml text section size (OTSS) due
to the stack overflow checks. We define OTSS as the sum of the sizes of all the
OCaml text sections in the compiled binary file ignoring the data sections, the
debug symbols, the text sections associated with OCaml runtime and other
statically linked C libraries. Compared to |stock|, OTSS is 19\% more for |MC|
and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
16-word red zone is effective at reducing OTSS compared to having no red zone,
whereas the 32-word red zone does not noticeably reduce OTSS further. Further
work is required to bring OTSS closer to |stock|.

@@

The results show that exceptions are no more expensive in |MC| compared to
|stock|. In the other programs, |MC| executes more instructions due to stack
overflow checks. The performance impact on callbacks is more significant than
external calls. For callbacks, since we reuse the current fiber stack, we need
to ensure it has enough room for inserting additional frames, while |stock|
does not need to do this. Callback performance is less important than external
calls, which are far more numerous. As seen previously, these results however
do not affect macro benchmark performance.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. These
programs do not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

@@

\begin{figure*}
	\includegraphics[angle=90,scale=0.44]{sandmark-notebook/sandmark_codesize}
	\caption{Normalized OCaml text section size (OTSS) of macro benchmarks.
	Baseline is Stock OCaml, whose size in bytes given in parenthesis.}
	\label{res:macro_size}
\end{figure*}

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
macro benchmarks that do not use effects using the Multicore OCaml compiler
which supports effect handlers. The aim is to quantify the impact of the
addition of effect handlers to the language on code that does not use effect
handlers. The |MC| is Multicore OCaml with the default red zone size (16
words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
overflow check. |MC+RedZone32| uses a red zone size of 32 words.
